\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()
# Load your packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)

# Set your ggplot theme
mytheme <- theme_set(theme_minimal())

# Load your datasets
PM2_5_2000 <- read_csv("DATA/PM2.5_2000.csv")
#View(PM2_5_2000)
PM2_5_2022 <- read_csv("DATA/PM2.5_2022.csv")
#View(PM2_5_2022)
SVI_CA_County_2000 <- read_csv("DATA/SVI_CA_County_2000.csv")
#View(SVI_CA_County_2000)
SVI_CA_County_2022 <- read_csv("DATA/SVI_CA_County_2022.csv")
#View(SVI_CA_County_2022)
```


# Rationale and Research Questions
This project explores the relationship between air quality and social vulnerability in California.

Research Question: Do communities facing higher pollution levels have less access to health insurance and medical care?

\newpage

# Dataset Information



\newpage

# Exploratory Analysis 
```{r}



```

\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
